### 2020-10-27 manuscript/manuscript.rmd
final pass to the manuscript and update figure and table orders

There has been considerable debate about the placement of the root in the animal tree of life, which has emerged as one of the most difficult problems in animal phylogenetics. This debate has major implications for our understanding of the earliest events in animal evolution, including the origin of the nervous system. Some phylogenetic analyses support a root that places the first split in the phylogeny of living animals between sponges and all other animals (the Porifera-sister hypothesis), and others find support for a split between comb jellies and all other animals (Ctenophora-sister). These analyses differ in many respects, including in the genes considered, species considered, molecular evolution models, and software. Here we systematically explore the rooting of the animal tree of life under consistent conditions by synthesizing data and results from 15 previous phylogenomic studies and performing a comprehensive set of new standardized analyses. It has previously been suggested that site-heterogeneous models favor Porifera-sister, but we find that this is not the case. Rather, Porifera-sister is only obtained under a narrow set of conditions when the number of site-heterogeneous categories is unconstrained and range into the hundreds. Site-heterogenous models with a fixed number of dozens of categories support Ctenophora-sister, and cross-validation indicates that such models fit the data just as well as the unconstrained models. Our analyses shed light on an important source of variation between phylogenomic analyses of the animal root. The datasets and analyses consolidated here will also be a useful test-platform for development of phylogenomic methods for this and other difficult problems.

@@

Historically, there was little debate about the root of the animal tree of life. Porifera-sister (Fig. 1E), the hypothesis that the animal root marks the divergence of Porifera (sponges) from all other animals (Fig. 1B), was widely accepted though rarely tested. By contrast, there has long been uncertainty about the placement of Ctenophora (comb jellies) (Fig. 1D) in the animal tree of life [@Wallberg:2004ws]. The first phylogenomic study to include ctenophores [@Dunn:2008ky] suggested a new hypothesis, now referred to as Ctenophora-sister, that ctenophores rather than sponges are our most distant living animal relative (Fig. 1A). Since then, many more phylogenomic studies have been published (Fig. 2), with some analyses finding support for Ctenophora-sister, some for Porifera-sister, and some neither [@King:2017ie] (Extended Data Table 1, Supplementary Information section S1). The extensive technical variation across these studies has been important to advancing our understanding of the sensitivity of these analyses, demonstrating for example that outgroup and model selection can have a large impact on these results (Fig. 3A). But the extensive technical variation has also made it difficult to synthesize these results to understand the underlying causes of this sensitivity. Several factors make resolution of the root of the animal tree a particularly challenging problem. For one, the nodes in question are the deepest in the animal tree of life. Another factor that has been invoked is branch lengths (*e.g.* Extended Data Figs. 1, 2), which are impacted by both divergence times and shifts in rate of evolution. Some sponges have a longer root to tip length, indicating an accelerated rate of evolution in those lineages. The stem branch of Ctenophora is longer than Porifera stem which, together with a more typical root-to-tip distance, is consistent with a more recent radiation of extant ctenophores [@podar2001] than extant poriferans. The longer ctenophore stem has led some to suggest that Ctenophora-sister could be an artefact of long-branch attraction to outgroups [@pisani2015genomic].

To advance this debate it is critical to examine variation in methods and results in a standardized and systematic framework. Here, we synthesize data and results from 15 previous phylogenomic studies that tested the Ctenophora-sister and Porifera-sister hypotheses (Fig. 2A, Extended Data Table 1). This set includes all phylogenomic studies of amino acid sequence data published before 2019 for which we could obtain data matrices with gene partition annotations. Among these 15 studies, three studies [@pisani2015genomic; @whelan2016let; @feuda2017improved] are based entirely on previously published data and gene-partition data is not available from one study [@Pick:2010eb].

@@

The models of sequence evolution in the studies considered here differ according to two primary components: the exchangeability matrix $R$ and amino acid equilibrium frequencies $\Pi$. The exchangeability matrix $R$ describes the relative rates at which one amino acid changes to another. The studies considered here use exchangeabilities that are the same between all amino acids (Poisson, also referred to as F81),or different. If different, the exchangeabilities can either be fixed based on previously empirically estimated rates (WAG or LG), or independently estimated from the data (GTR). The analyses considered here have site-homogeneous exchangeability models (site-homogeneous model), which means that the same matrix is used for all sites. The equilibrium frequencies describe the expected frequency of each amino acid, which captures the fact that some amino acids are much more common than others. The published analyses differ in whether they take a homogeneous approach and jointly estimate the same frequency across all sites in a partition, or add parameters that allow heterogeneous equilibrium frequencies that differ across sites. Heterogeneous approaches include CAT [@Lartillot:2004dq], which is implemented in the software PhyloBayes and has been widely applied to questions of deep animal phylogeny. The models that are applied in practice are heavily influenced by computational costs, model availability in software, and convention. While studies often discuss CAT and WAG models as if they are mutually exclusive, we note that these particular terms apply to non-exclusive model components -- CAT refers to heterogeneous equilibrium frequencies across sites and WAG to a particular exchangeability matrix. In this literature, CAT is generally shorthand for Poisson+CAT and WAG is shorthand for WAG+homogeneous equilibrium frequency estimation. To avoid confusion on this point, here we always specify the exchangeability matrix first (*e.g.* GTR), followed by modifiers that describe the accommodation of heterogeneity in equilibrium frequencies (*e.g.* CAT). If site homogeneous equilibrium frequencies are used, we refer to the exchangeability matrix alone. Gamma-rate heterogeneity, a scalar that accommodates total rate of change across sites, is used in almost every analysis conducted here and we generally omit its designation. Some analyses partition the data by genes, and use different models for each gene (Supplementary Information section S2).

High-throughput sequencing allows investigators to readily assemble matrices with hundreds or thousands of protein-coding genes from a broad diversity of animal species (Extended Data Table 1). Studies of animal phylogeny have used a wide variety of different approaches to identifying and selecting genes and taxa for their matrices. As a result, the particular genes selected for analysis vary widely (Fig. 2B, Extended Data Fig. 4). Gene sampling varies in several ways, including in the fractions of single copy orthologs (*e.g.* BUSCO genes) and ribosomal protein genes in the matrix (Extended Data Fig. 3).

@@

Ingroup taxon sampling also varies widely between studies (Fig. 2B, Extended Data Fig. 4). Sampling of ingroup taxa (animals) in early studies was biased toward Bilateria. Sampling of non-bilaterian animals, including sponges and ctenophores, has improved over time (Extended Data Fig. 4). Within each clade, there is often considerable variation in taxon sampling and therefore often little species overlap across studies (Extended Data Fig. 4). This variation is in part because newer sequencing technologies in more recent studies are usually not applied to the exact same species that were included in earlier studies.

Sampling of outgroup taxa (non-animals, in this case) is critical to phylogenetic rooting questions, since the node where the outgroup subtree attaches to the ingroup subtree is the root of the ingroup. There has therefore been extensive focus on improving outgroup sampling when testing phylogenetic hypotheses about rooting [@graham2002rooting]. Most studies addressing the animal root have removed more distantly related outgroup taxa in some analyses to explore the effect of outgroup selection to ingroup topology [@ryan2013genome; @pisani2015genomic]. Three progressively more inclusive clades have often been investigated: Choanimalia (animals plus most closely related Choanoflagellatea), Holozoa (Choanimalia plus more distantly related Holozoa) and Opisthokonta (Holozoa + Fungi).

@@

![](figures/Figure3_published.png)
**Fig. 3.**  Summary of phylogenomic results from previous studies and reanalysis conducted in this study. Each point represents a phylogenetic analysis with a support of bootstrap values less than `r bootstrap_threshold`% or a posterior probability less than `r posterior_prob_threshold`% are considered as unresolved. The clades are organized with increased outgroup sampling higher on the vertical axis, and the models are organized in general by increasing complexity in terms of numbers of parameters to the right on the horizontal axis. (A) Analyses transcribed from the literature (related to Supplementary Table 2). (B) New phylogenomic analyses conducted in this study (related to Supplementary Table 3).

We parsed `r analyses_processed %>% filter(status=="published") %>% nrow()` previous phylogenetic analyses from `r analyses_processed %>% filter(status=="published") %>% pull (manuscript) %>% unique() %>% length()` studies (Fig. 3A and Supplementary Table 2). The conclusions of five studies strongly favor Porifera-sister and 10 favor Ctenophora-sister (Extended Data Table 1). Three studies are based entirely on previously published data, and the remainder add data for one or more species.

@@

Our summary of previous phylogenetic analyses (Fig. 3A) shows that Ctenophora-sister is supported in analyses that span the full range of outgroup sampling and models used to date, including some analyses with restricted outgroup sampling and models that accommodate site-heterogeneous equilibrium frequencies with CAT. This is consistent with previous assessments of the problem [@whelan2016let], but is drawn from a much more extensive and systematic examination. The only analyses that support Porifera-sister have a reduced outgroup sampling (Choanimalia, Holozoa) and site-heterogeneous models with CAT. Model adequacy assessments generally favor GTR+CAT over Poisson+CAT or site-homogeneous models [@pisani2015genomic; @feuda2017improved], but because GTR+CAT is so parameter rich, many analyses that use a model with GTR+CAT do not converge. The fact that Porifera-sister is recovered only for particular models with particular outgroup sampling indicates that model and outgroup interact, and that this interaction is fundamental to understanding the range of results obtained across analyses.

@@

For all but the three largest matrices, we tested the relative fit of a variety of models, both with and without C10-C60 accommodation of site heterogeneity in equilibrium frequencies, using ModelFinder [@kalyaanamoorthy2017modelfinder] in IQ-TREE. In all cases, models with C60 fit these matrices better than the site-homogeneous models. This is consistent with the importance of accommodating site heterogeneity noted by previous investigators [@Lartillot:2004dq; @Philippe:2009hh; @nosenko2013deep; @pisani2015genomic; @simion2017large; @feuda2017improved]. We then inferred support under the best-fit model (Supplementary Table 1), except for the three largest matrices where we used LG+C60. We then analyzed each matrix under a panel of standard site-heterogeneous and site-homogeneous models, including WAG, GTR and Poisson+C60 (Supplementary Table 1).

All IQ-TREE analyses, apart from unresolved analyses (for Moroz2014_3d and all Nosenko2013 matrices), supported Ctenophora-sister (Fig. 3B). No IQ-TREE analyses supported Porifera-sister, including those that restrict outgroup sampling to Choanoflagellata and use models with site-heterogeneous equilibrium frequencies (Fig. 3B lower right; Supplementary Table 3), the conditions under which published PhyloBayes CAT analyses recover strong support for Porifera-sister (Fig. 3A). To further verify this difference in a controlled manner we reran PhyloBayes analyses with CAT, using both Poisson and GTR substitution matrices, for some matrices that had led to support for Porifera-sister in published analyses. Consistent with published results, some of these supported Porifera-sister.

@@

We examined the specific impact of this large difference in category number on animal root position. It is currently not possible to use more than 60 categories for C models in IQ-TREE, but the number of categories can be set *a priori* in Phylobayes using nCAT. We therefore varied the number of categories in Phylobayes analyses (Fig. 4; Supplementary Table 6). We found that Poisson+nCAT60 analyses in PhyloBayes, like Poisson+C60 and WAG+C60 IQ-TREE analyses, provide strong support of Ctenophora-sister. This indicates that the difference in results between unconstrained CAT analyses in PhyloBayes and C60 analyses in IQ-TREE is not due to differences in the software or other implementation factors, but due to the large difference, in excess of ten-fold, in the number of site categories. When we increased the number of categories in PhyloBayes nCAT analyses, we observe the transition from support for Ctenophora-sister to an unresolved root to Porifera-sister (Fig. 4). For example, for the Whelan2017_strict matrix this transition occurs between 60 -- 120 categories when using Poisson+nCAT model. Due to computational limitations of GTR models, we only ran GTR+CAT and GTR+nCAT60 models on representative matrices with Choanimalia sampling. We found that for Whelan2017 matrices, support shifted from Porifera-sister with Poisson+CAT model to Ctenophora-sister using GTR+CAT model. Moreover, we also found all results strongly supported Ctenophora-sister with GTR+nCAT60 models.

@@

The first questions regard the relative fit of adding aditional site-heterogeneous categories, and the adequacy of models with different numbers of categories. ModelFinder selects site heterogeneous C60 models (Supplementary Table 5), but IQtree often gives a warning under C60 that the model may be overfit, with too many categories, because some mixture weights are close to 0. While ModelFinder considers the relative fit of models, this is distinct from model adequacy -- an absolute measure of how well the model can describe the variation in the data. Cross Validation (CV) is an important tool to assess adequacy by examining the predictive power of the model, *i.e.* the ability of the model to predict observations [@lartillot2009phylobayes]. It could be that a model with a very large number of categories are required to adequately describe the site heterogeneity in these data, and that we should therefore favor the Porifera-sister reseults obtained in unconstrained CAT analyses. To evaluate the predictive performance and model-fit between Poisson+CAT and Poisson+nCAT60 models in Whelan2017_strict and Philippe2009_Choanimalia matrices, we conducted Bayesian cross-validation experiments with PhyloBayes. We found nearly identical distributions of cross validation scores for Poisson+nCAT60 and Poisson+CAT models (Extended Data Fig. 6). A t-test analysis shows that there is no significant difference between them for either matrix (Whelan2017_strict: $p-value=`r P_value_Whelan2017_strict`$; Philippe2009_Choanimalia: $p-value=`r P_value_Philippe2009`$). This indicates that 60 categories in a Poisson+nCAT60 model are as as adequate at predicting the data as the hundreds of categories in an unconstrained Poisson+CAT model. This suggests there is no reason to favor these narrow analysis conditions under which Porifera-sister are obtained.

In a site-heterogeneous model, each site is allocated to one of the categories in each sample. Different categories can have very different numbers of sites. We examine category allocations in the last chain samples of analyses of the Philippe2009_Choanimalia and Whelan2017_strict matrices (Extended Data Fig. 5A-B), which have `r n_categories_phil_cat` and `r n_categories_whel_cat` categories, repectively. The fraction of sites allocated to the 50% of the least frequent categories is `r mid_phil_cat`% in the analysis of the Philippe2009_Choanimalia matrix and `r mid_whel_cat`% in the analysis of the Whelan2017_strict matrix. This long tail of categories that apply to a very small fraction of sites is in stark contrast to nCAT60 analyses, constrained to 60 categories, which have no such long tail of rare categories (Extended Data Fig. 5A-B). The fact that so many categories apply to such a small fraction of data be help explain why increasing the number of categories more than ten-fold has so little impact on predictive power of these far more complex models.

@@

```
![](figures/Figure4_manuscript.png)
**Fig. 4.**  Sensitivity analyses with representative data matrices analyzed by different number of equilibrium frequency categories (nCAT) in PhyloBayes. Statistical support values (posterior probabilities) were obtained from three data matrices using the site-heterogeneous Poisson+CAT model with different categories. (A) Phlippe2009_Choanozoa; (B) Philippe2009_Holozoa; (C) Whelan2017_full_Choanozoa; (D) Whelan2017_strict. Statistical support for Ctenophora-sister and Porifera-sister is indicated in orange and green, respectively. Support values from the sensitive analyses are shown in Supplementary Table 6.

@@

To further explore the phylogenetic signal of different models we discussed above, we quantified the phylogenetic signal for Porifera-sister and Ctenophora-sister topologies across three representative data matrices when varying outgroup sampling and model (Extended Data Fig. 7). By calculating differences in log-likelihood scores for these topologies for every gene ($\Delta$|lnL|) in each matrixin each matrix when using site-homogeneous models in IQ-TREE, we found that the Ctenophora-sister had the higher proportions of supporting genes in every analysis. Moreover, outgroup choice has little impact on the distribution of the support for phylogenetic signals in analyses with site-homogeneous models. This finding is largely consistent with the previously observed distribution of support for Ctenophora-sister in other data matrices [@shen2017contentious].

@@

We constructed a network with Python and NetworkX v2.2 [@Hagberg:2008] where each node is a partition and each edge represents a DIAMOND sequence-to-sequence match between sequences in the partitions. We extracted each connected component from this network. We further split these components if the the most connected node (i.e. most edges) had two times more the standard deviation from the mean number of edges in the component it is a member of and if removing that node splits the component into two or more components. We then decorated every node in the partition network with the most often found SwissProt BLAST+ result and BUSCO results to see which components contain which classes and families of genes. See partition_network_summary in Rdata for a summary tally of each part of the comparison.

@@

#### Outgroup taxa sampling with C60 and CAT models
Because different choices of outgroups could also affect phylogenetic inference as suggested in previous analyses, we parsed the full data matrices into three different types of outgroups: Choanimalia, Holozoa and Opisthokonta. These datasets include the same set of genes but differ in the composition of outgroup species. Choanimalia only includes choanofagellates as outgroup; Holozoa also includes more distantly related Holozoans; Opistokonta also includes Fungi. For each Choanimalia data matrice, both C60 models in IQ-TREE and Poisson+CAT models in PhyloBayes were conducted. The maximum likelihood analysis was performed using the best-fit substitution model identified as above and nodal support was assessed with 1000 ultrafast bootstrap replicates using IQ-TREE. Moreover, bayesian inference with the site-heterogeneous Poisson+CAT model was done with PhyloBayes-MPI v1.8. To minimize computational burden, GTR+CAT models were only performed in the representative Choanimalia matrices from Philippe2009, Ryan2013_est and Whelan2017_full.

For several Choanozoa matrices indicated strong support for the hypothesis that sponges are the sister group to the remaining Metazoa using the Poisson+CAT model, bayesian inference with Poisson+CAT model was also conducted to Holozoa and Opisthokonta data matrices with the same settings as above. For all the analyses with Poisson+CAT models in PhyloBayes, two independent chains were sampled every generation. Tracer plots of MCMC runs were visually inspected in Tracer v1.6 to assess stationarity and appropriate burn-in. Chains were considered to have reached convergence when the maxdiff statistic among chains was below 0.3 (as measured by bpcomp) and effective sample size > 50 for each parameter (as measured by tracecomp). A 50% majority-rule consensus tree was computed with bpcomp, and nodal support was estimated by posterior probability. Most Phylobayes runs converged, although several large matrices have not reached convergence after at least a month's computational time. For those matrices that were not converged, PhyloBayes analyses were run for at least two weeks. We also summarized the average number of substitutional categories inferred for each PhyloBayes analysis using Tracer.

@@

## Supplementary tables

>>>>

There has been considerable debate about the placement of the root in the animal tree of life, which has emerged as one of the most challenging problems in animal phylogenetics. This debate has major implications for our understanding of the earliest events in animal evolution, including the origin of the nervous system. Some phylogenetic analyses support a root that places the first split in the phylogeny of living animals between sponges and all other animals (the Porifera-sister hypothesis), and others find support for a split between comb jellies and all other animals (Ctenophora-sister). These analyses differ in many respects, including in the genes considered, species considered, molecular evolution models, and software. Here we systematically explore the rooting of the animal tree of life under consistent conditions by synthesizing data and results from 15 previous phylogenomic studies and performing a comprehensive set of new standardized analyses. It has previously been suggested that site-heterogeneous models favor Porifera-sister, but we find that this is not the case. Rather, Porifera-sister is only obtained under a narrow set of conditions when the number of site-heterogeneous categories is unconstrained and range into the hundreds. Site-heterogenous models with a fixed number of dozens of categories support Ctenophora-sister, and cross-validation indicates that such models fit the data just as well as the unconstrained models. Our analyses shed light on an important source of variation between phylogenomic studies of the animal root. The datasets and analyses consolidated here will also be a useful test-platform for the development of phylogenomic methods for this and other difficult problems.

@@

Historically, there was little debate about the root of the animal tree of life. Porifera-sister (Fig. 1E), the hypothesis that the animal root marks the divergence of Porifera (sponges) from all other animals (Fig. 1B), was widely accepted though rarely tested. By contrast, there has long been uncertainty about the placement of Ctenophora (comb jellies) (Fig. 1D) in the animal tree of life [@Wallberg:2004ws]. The first phylogenomic study to include ctenophores [@Dunn:2008ky] suggested a new hypothesis, now referred to as Ctenophora-sister, that ctenophores rather than sponges are our most distant living animal relative (Fig. 1A). Since then, many more phylogenomic studies have been published (Fig. 2), with some analyses finding support for Ctenophora-sister, some for Porifera-sister, and some neither [@King:2017ie] (Extended Data Table 1, Supplementary Information section S1). The extensive technical variation across these studies has been important to advancing our understanding of the sensitivity of these analyses, demonstrating for example that outgroup and model selection can have a large impact on these results (Fig. 3A). But the extensive technical variation has also made it difficult to synthesize these results to understand the underlying causes of this sensitivity. Several factors make resolution of the root of the animal tree a particularly challenging problem. For one, the nodes in question are the deepest in the animal tree of life. Another factor that has been invoked is branch lengths (*e.g.* Extended Data Figs. 1, 2), which are impacted by both divergence times and shifts in rate of evolution. Some sponges have a longer root to tip length, indicating an accelerated rate of evolution in those lineages. The stem branch of Ctenophora is longer than Porifera stem which, together with a more typical root-to-tip distance, is consistent with a more recent radiation of extant ctenophores [@podar2001] than extant poriferans. The longer ctenophore stem has led some to suggest that Ctenophora-sister could be an artifact of long-branch attraction to outgroups [@pisani2015genomic].

To advance this debate it is critical to examine variation in methods and results in a standardized and systematic framework. Here, we synthesize data and results from 15 previous phylogenomic studies that tested the Ctenophora-sister and Porifera-sister hypotheses (Fig. 2, Extended Data Table 1). This set includes all phylogenomic studies of amino acid sequence data published before 2019 for which we could obtain data matrices with gene partition annotations. Among these 15 studies, three studies [@pisani2015genomic; @whelan2016let; @feuda2017improved] are based entirely on previously published data, and gene-partition data is not available from one study [@Pick:2010eb].

@@

The models of sequence evolution in the studies considered here differ according to two primary components: the exchangeability matrix $R$ and amino acid equilibrium frequencies $\Pi$. The exchangeability matrix $R$ describes the relative rates at which one amino acid changes to another. The studies considered here use exchangeabilities that are the same between all amino acids (Poisson, also referred to as F81), or different. If different, the exchangeabilities can either be fixed based on previously empirically estimated rates (WAG or LG), or independently estimated from the data (GTR). The analyses considered here have site-homogeneous exchangeability models (site-homogeneous model), which means that the same matrix is used for all sites. The equilibrium frequencies describe the expected frequency of each amino acid, which captures the fact that some amino acids are much more common than others. The published analyses differ in whether they take a homogeneous approach and jointly estimate the same frequency across all sites in a partition, or add parameters that allow heterogeneous equilibrium frequencies that differ across sites. Heterogeneous approaches include CAT [@Lartillot:2004dq], which is implemented in the software PhyloBayes and has been widely applied to questions of deep animal phylogeny. The models that are applied in practice are heavily influenced by computational costs, model availability in software, and convention. While studies often discuss CAT and WAG models as if they are mutually exclusive, we note that these particular terms apply to non-exclusive model components -- CAT refers to heterogeneous equilibrium frequencies across sites and WAG to a particular exchangeability matrix. In this literature, CAT is generally shorthand for Poisson+CAT and WAG is shorthand for WAG+homogeneous equilibrium frequency estimation. To avoid confusion on this point, here we always specify the exchangeability matrix first (*e.g.* GTR), followed by modifiers that describe the accommodation of heterogeneity in equilibrium frequencies (*e.g.* CAT). If site homogeneous equilibrium frequencies are used, we refer to the exchangeability matrix alone. Gamma-rate heterogeneity, a scalar that accommodates the total rate of change across sites, is used in almost every analysis conducted here and we generally omit its designation. Some analyses partition the data by genes, and use different models for each gene (Supplementary Information section S2).

High-throughput sequencing allows investigators to readily assemble matrices with hundreds or thousands of protein-coding genes from a broad diversity of animal species (Extended Data Table 1). Studies of animal phylogeny have used a wide variety of different approaches to identifying and selecting genes and taxa for their matrices. As a result, the particular genes selected for analysis vary widely (Fig. 2, Extended Data Fig. 3). Gene sampling varies in several ways, including in the fractions of single-copy orthologs (*e.g.* BUSCO genes) and ribosomal protein genes in the matrix (Extended Data Fig. 4).

@@

Ingroup taxon sampling also varies widely between studies (Fig. 2, Extended Data Fig. 3). Sampling of ingroup taxa (animals) in early studies was biased toward Bilateria. Sampling of non-bilaterian animals, including sponges and ctenophores, has improved over time (Extended Data Fig. 3). Within each clade, there is often considerable variation in taxon sampling and therefore often little species overlap across studies (Extended Data Fig. 3). This variation is in part because newer sequencing technologies in more recent studies are usually not applied to the exact same species that were included in earlier studies.

Sampling of outgroup taxa (non-animals, in this case) is critical to phylogenetic rooting questions, since the node where the outgroup subtree attaches to the ingroup subtree is the root of the ingroup. There has therefore been extensive focus on improving outgroup sampling when testing phylogenetic hypotheses about rooting [@graham2002rooting]. Most studies addressing the animal root have removed more distantly related outgroup taxa in some analyses to explore the effect of outgroup selection to ingroup topology [@ryan2013genome; @pisani2015genomic]. Three progressively more inclusive clades have often been investigated: Choanimalia (animals plus most closely related Choanoflagellatea), Holozoa (Choanimalia plus more distantly related Holozoa), and Opisthokonta (Holozoa + Fungi).

@@

![](figures/Figure3_published.png)
**Fig. 3.**  Summary of phylogenomic results from previous studies and reanalysis conducted in this study. Each point represents a phylogenetic analysis with a support of bootstrap values less than `r bootstrap_threshold`% or a posterior probability less than `r posterior_prob_threshold`% are considered as unresolved. The clades are organized with increased outgroup sampling higher on the vertical axis, and the models are organized in general by increasing complexity in terms of numbers of parameters to the right on the horizontal axis. (A) Analyses transcribed from the literature (related to Supplementary Table 1). (B) New phylogenomic analyses conducted in this study (related to Supplementary Table 2). Over 1,011,765 CPU hours (~115 CPU years) were used for this study.

We parsed `r analyses_processed %>% filter(status=="published") %>% nrow()` previous phylogenetic analyses from `r analyses_processed %>% filter(status=="published") %>% pull (manuscript) %>% unique() %>% length()` studies (Fig. 3A and Supplementary Table 1). The conclusions of five studies strongly favor Porifera-sister and ten favor Ctenophora-sister (Extended Data Table 1). Three studies are based entirely on previously published data, and the remainder add data for one or more species.s

@@

OOur summary of previous phylogenetic analyses (Fig. 3A) shows that Ctenophora-sister is supported in analyses that span the full range of outgroup sampling and models used to date, including some analyses with restricted outgroup sampling and models that accommodate site-heterogeneous equilibrium frequencies with CAT. This is consistent with previous assessments of the problem [@whelan2016let], but is drawn from a much more extensive and systematic examination. The only analyses that support Porifera-sister have reduced outgroup sampling (Choanimalia, Holozoa) and site-heterogeneous models with CAT. Model adequacy assessments generally favor GTR+CAT over Poisson+CAT or site-homogeneous models [@pisani2015genomic; @feuda2017improved], but because GTR+CAT is so parameter rich, many analyses that use a model with GTR+CAT do not converge. The fact that Porifera-sister is recovered only for particular models with particular outgroup sampling indicates that model and outgroup interact, and that this interaction is fundamental to understanding the range of results obtained across analyses.

@@

For all but the three largest matrices, we tested the relative fit of a variety of models, both with and without C10-C60 accommodation of site heterogeneity in equilibrium frequencies, using ModelFinder [@kalyaanamoorthy2017modelfinder] in IQ-TREE. In all cases, models with C60 fit these matrices better than the site-homogeneous models. This is consistent with the importance of accommodating site heterogeneity noted by previous investigators [@Lartillot:2004dq; @Philippe:2009hh; @nosenko2013deep; @pisani2015genomic; @simion2017large; @feuda2017improved]. We then inferred support under the best-fit model (Supplementary Table 3), except for the three largest matrices where we used LG+C60. We then analyzed each matrix under a panel of standard site-heterogeneous and site-homogeneous models, including WAG, GTR and Poisson+C60 (Supplementary Table 3).

All IQ-TREE analyses, apart from unresolved analyses (for Moroz2014_3d and all Nosenko2013 matrices), supported Ctenophora-sister (Fig. 3B). No IQ-TREE analyses supported Porifera-sister, including those that restrict outgroup sampling to Choanoflagellata and use models with site-heterogeneous equilibrium frequencies (Fig. 3B lower right; Supplementary Table 2), the conditions under which published PhyloBayes CAT analyses recover strong support for Porifera-sister (Fig. 3A). To further verify this difference in a controlled manner we reran PhyloBayes analyses with CAT, using both Poisson and GTR substitution matrices, for some matrices that had led to support for Porifera-sister in published analyses. Consistent with published results, some of these supported Porifera-sister.

@@

We examined the specific impact of this large difference in category number on animal root position. It is currently not possible to use more than 60 categories for C models in IQ-TREE, but the number of categories can be set *a priori* in Phylobayes using nCAT. We therefore varied the number of categories in Phylobayes analyses (Fig. 4; Supplementary Table 5). We found that Poisson+nCAT60 analyses in PhyloBayes, like Poisson+C60 and WAG+C60 IQ-TREE analyses, provide strong support of Ctenophora-sister. This indicates that the difference in results between unconstrained CAT analyses in PhyloBayes and C60 analyses in IQ-TREE is not due to differences in the software or other implementation factors, but due to the large difference, in excess of ten-fold, in the number of site categories. When we increased the number of categories in PhyloBayes nCAT analyses, we observe the transition from support for Ctenophora-sister to an unresolved root to Porifera-sister (Fig. 4). For example, for the Whelan2017_strict matrix this transition occurs between 60 -- 120 categories when using Poisson+nCAT model. Due to computational limitations of GTR models, we only ran GTR+CAT and GTR+nCAT60 models on representative matrices with Choanimalia sampling. We found that for Whelan2017 matrices, support shifted from Porifera-sister with Poisson+CAT model to Ctenophora-sister using GTR+CAT model. Moreover, we also found all results strongly supported Ctenophora-sister with GTR+nCAT60 models.

@@

The first questions regard the relative fit of adding aditional site-heterogeneous categories, and the adequacy of models with different numbers of categories. ModelFinder selects site heterogeneous C60 models (Supplementary Table 3), but IQ-TREE often gives a warning under C60 that the model may overfit, with too many categories, because some mixture weights are close to 0 (Supplementary Table 6). While ModelFinder considers the relative fit of models, this is distinct from model adequacy -- an absolute measure of how well the model can describe the variation in the data. Cross-Validation (CV) is an important tool to assess adequacy by examining the predictive power of the model, *i.e.* the ability of the model to predict observations [@lartillot2009phylobayes]. It could be that a model with a very large number of categories are required to adequately describe the site heterogeneity in these data, and that we should therefore favor the Porifera-sister results obtained in unconstrained CAT analyses. To evaluate the predictive performance and model-fit between Poisson+CAT and Poisson+nCAT60 models in Whelan2017_strict and Philippe2009_Choanimalia matrices, we conducted Bayesian cross-validation experiments with PhyloBayes. We found nearly identical distributions of cross-validation scores for Poisson+nCAT60 and Poisson+CAT models (Extended Data Fig. 5). A t-test analysis shows that there is no significant difference between them for either matrix (Whelan2017_strict: $p-value=`r P_value_Whelan2017_strict`$; Philippe2009_Choanimalia: $p-value=`r P_value_Philippe2009`$). This indicates that 60 categories in a Poisson+nCAT60 model are as adequate as at predicting the data as the hundreds of categories in an unconstrained Poisson+CAT model. This suggests there is no reason to favor these narrow analysis conditions under which Porifera-sister are obtained.

In a site-heterogeneous model, each site is allocated to one of the categories in each sample. Different categories can have very different numbers of sites. We examine category allocations in the last chain samples of analyses of the Philippe2009_Choanimalia and Whelan2017_strict matrices (Extended Data Fig. 6A-B), which have `r n_categories_phil_cat` and `r n_categories_whel_cat` categories, respectively. The fraction of sites allocated to the 50% of the least frequent categories is `r mid_phil_cat`% in the analysis of the Philippe2009_Choanimalia matrix and `r mid_whel_cat`% in the analysis of the Whelan2017_strict matrix. This long tail of categories that apply to a very small fraction of sites is in stark contrast to nCAT60 analyses, constrained to 60 categories, which have no such long tail of rare categories (Extended Data Fig. 6A-B). The fact that so many categories apply to such a small fraction of data be help explain why increasing the number of categories more than ten-fold has so little impact on the predictive power of these far more complex models.

@@

```
![](figures/Figure4_manuscript.png)
**Fig. 4.**  Sensitivity analyses with representative data matrices were analyzed by different number of equilibrium frequency categories (nCAT) in PhyloBayes. Statistical support values (posterior probabilities) were obtained from three data matrices using the site-heterogeneous Poisson+CAT model with different categories. (A) Phlippe2009_Choanozoa; (B) Philippe2009_Holozoa; (C) Whelan2017_full_Choanozoa; (D) Whelan2017_strict. Statistical support for Ctenophora-sister and Porifera-sister is indicated in orange and green, respectively. Support values from the sensitivity analyses are shown in Supplementary Table 5.

@@

To further explore the phylogenetic signal of different models we discussed above, we quantified the phylogenetic signal for Porifera-sister and Ctenophora-sister topologies across three representative data matrices when varying outgroup sampling and model (Extended Data Fig. 7). By calculating differences in log-likelihood scores for these topologies for every gene ($\Delta$|lnL|) in each matrix when using site-homogeneous models in IQ-TREE, we found that the Ctenophora-sister had the higher proportions of supporting genes in every analysis. Moreover, outgroup choice has little impact on the distribution of the support for phylogenetic signals in analyses with site-homogeneous models. This finding is largely consistent with the previously observed distribution of support for Ctenophora-sister in other data matrices [@shen2017contentious].

@@

We constructed a network with Python and NetworkX v2.2 [@Hagberg:2008] where each node is a partition and each edge represents a DIAMOND sequence-to-sequence match between sequences in the partitions. We extracted each connected component from this network. We further split these components if the most connected node (i.e. most edges) had two times more the standard deviation from the mean number of edges in the component it is a member of and if removing that node splits the component into two or more components. We then decorated every node in the partition network with the most often found SwissProt BLAST+ result and BUSCO results to see which components contain which classes and families of genes. See partition_network_summary in Rdata for a summary tally of each part of the comparison.

@@

#### Outgroup taxa sampling with C60 and CAT models
Because different choices of outgroups could also affect phylogenetic inference as suggested in previous analyses, we parsed the full data matrices into three different types of outgroups: Choanimalia, Holozoa and Opisthokonta. These datasets include the same set of genes but differ in the composition of outgroup species. Choanimalia only includes choanofagellates as outgroup; Holozoa also includes more distantly related Holozoans; Opistokonta also includes Fungi. For each Choanimalia data matrice, both C60 models in IQ-TREE and Poisson+CAT models in PhyloBayes were conducted. The maximum likelihood analysis was performed using the best-fit substitution model identified as above and nodal support was assessed with 1000 ultrafast bootstrap replicates using IQ-TREE. Moreover, Bayesian inference with the site-heterogeneous Poisson+CAT model was done with PhyloBayes-MPI v1.8. To minimize computational burden, GTR+CAT models were only performed in the representative Choanimalia matrices from Philippe2009, Ryan2013_est and Whelan2017_full.

For several Choanozoa matrices indicated strong support for the hypothesis that sponges are the sister group to the remaining Metazoa using the Poisson+CAT model, Bayesian inference with Poisson+CAT model was also conducted to Holozoa and Opisthokonta data matrices with the same settings as above. For all the analyses with Poisson+CAT models in PhyloBayes, two independent chains were sampled every generation. Tracer plots of MCMC runs were visually inspected in Tracer v1.6 to assess stationarity and appropriate burn-in. Chains were considered to have reached convergence when the maxdiff statistic among chains was below 0.3 (as measured by bpcomp) and effective sample size > 50 for each parameter (as measured by tracecomp). A 50% majority-rule consensus tree was computed with bpcomp, and nodal support was estimated by posterior probability. Most Phylobayes runs converged, although several large matrices have not reached convergence after at least a month's computational time. For those matrices that were not converged, PhyloBayes analyses were run for at least two weeks. We also summarized the average number of substitutional categories inferred for each PhyloBayes analysis using Tracer.

@@
